---
title: "fastq"
author: "nati piland"
date: "2019-09-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


```{r}
library(tidyverse)
#like the tree command on terminal
new_files <- dir("data", 
                  recursive=TRUE, 
                  full.names=TRUE)
```

To subset the data use the `[` operator. To find a string use the `grepl` function.

```{r}
fastq_files <- new_files[grepl("fastq",new_files)]
fastq_1 <- read_lines(fastq_files[1])
fastq_2 <- read_lines(fastq_files[2])
head(fastq_1)
```

Fastq files have four lines per read. 
1) header line has metadata
2) sequence
3) same metadata?
4) quality score per base

```{r}
fastq_fields <- c("metadata",
                 "sequence",
                 "metadata2",
                 "score")
length(fastq_1)
#This part helps put an ID to all the reads in the fastq dataframe so that it can spread properly.
num_reads <- length(fastq_1)/4
read_id <- gl(num_reads, k=4)
fastq_fields <- rep_len(fastq_fields, 
                        length.out = length(fastq_1))
length(fastq_fields)
```

## Creating Dataframes (`tibbles`)

```{r}
fastq_df <- tibble(read = read_id,
                   field = fastq_fields,
                   lines = fastq_1)
fastq_df <- fastq_df %>% 
  filter(field != "metadata2")

pivot_fastq_df <- fastq_df %>%
  spread(key = field, 
         value = lines)
```

```{r}
head(pivot_fastq_df) %>% 
  pull(metadata)
#Take out the information from metadata into three different columns, and change length into an integer.
pivot_fastq_df <- separate(pivot_fastq_df, 
                           metadata, 
                           into = c("sr_id", 
                                    "tag", 
                                    "length"), 
                           sep = " ")

pivot_fastq_df <- mutate(pivot_fastq_df, 
                         length = str_replace(length, 
                                              pattern = "length=",
                                              replacement = "")) %>%
                  mutate(length=as.integer(length))
#Make sure that all rows are 36
filter(pivot_fastq_df, length != 36)
```


## Refactoring

Refactoring is changing code without changing its behavior. You do this because certain code is easier to change than other code: You're rarely done when you think you're done. A powerful way to do this is to take all your code and turn it into a function. 

Let's refactor our code to be a function that takes a fastq file and turns it into a data frame.

```{r}
read_fastq <- function(file,name=file){
  fastq_1 <- read_lines(file)
  fastq_fields <- c("metadata",
                 "sequence",
                 "metadata2",
                 "score")
  num_reads <- length(fastq_1)/4
  read_id <- gl(num_reads, k=4)
  fastq_fields <- rep_len(fastq_fields, 
                        length.out = length(fastq_1))
  fastq_df <- tibble(name = name,
                   read = read_id,
                   field = fastq_fields,
                   lines = fastq_1)
  fastq_df <- fastq_df %>% 
                  filter(field != "metadata2")
  pivot_fastq_df <- fastq_df %>%
                        spread(key = field, 
                        value = lines)
  pivot_fastq_df <- separate(pivot_fastq_df, 
                           metadata, 
                           into = c("sr_id", 
                                    "tag", 
                                    "length"), 
                           sep = " ")

pivot_fastq_df <- mutate(pivot_fastq_df, 
                         length = str_replace(length, 
                                              pattern = "length=",
                                              replacement = "")) %>%
                  mutate(length=as.integer(length))
return(pivot_fastq_df)
}
```

```{r}
fastq_df_1 <- read_fastq(fastq_files[1], name = "SRR097977")
fastq_df_2 <- read_fastq(fastq_files[2], name = "SRR098026")

#If you already have the dataframes, bind them.
fastq_df <- bind_rows(fastq_df_1,
                      fastq_df_2)

#A fancier way to do this is:
fastq_df <- map_df(fastq_files, 
                   read_fastq)
fastq_df <- map2_df(fastq_files, 
                    c("SRR097977", "SRR098026"), 
                    read_fastq)
```

The number of rows across fastq files that have an N in the sequence.

```{r}
#Three ways to answer this.
fastq_df_n <- fastq_df[grepl("N", fastq_df$sequence),]
fastq_df_n <- fastq_df %>%
                filter(grepl("N", fastq_df$sequence))
#For these last two, just look at the dimensions of the data. Or:
fastq_df %>%
  filter(grepl("N", fastq_df$sequence)) %>%
  group_by(name) %>%
  summarise(num_N=n())

#Creating lists out of the sequences. As fundamental of a data type as vectors. 
fastq_df <- fastq_df %>%
  mutate(seq_list=str_split(sequence, 
                            pattern = ""))

fastq_seq_list <- fastq_df$seq_list
#This is to get the first element of the list. This is the samd as fastq_df$sequence[2], except that seq_list gives you a vector of strings.
fastq_seq_list[[1]]

count_N <- function(x){
  sum(x=='N')
}
count_N(fastq_seq_list[[1]])
map_int(fastq_seq_list, count_N)
fastq_df <- fastq_df %>%
              mutate(num_N=map_int(seq_list, count_N))

ggplot(fastq_df, aes(x=num_N)) +
  geom_histogram()
```

How many of the question mark scores are there at every read?

```{r}
#Creating lists out of the sequences. As fundamental of a data type as vectors. 
fastq_df <- fastq_df %>%
  mutate(score_list=str_split(score, 
                            pattern = ""))

fastq_score_list <- fastq_df$score_list
#This is to get the first element of the list. This is the samd as fastq_df$sequence[2], except that seq_list gives you a vector of strings.
fastq_score_list[[1]]

count_q <- function(x){
  sum(x=='?')
}
#could also straight up make this function more generic
count_char <- function(x,char="?"){
  sum(x==char)
}

count_q(fastq_score_list[[1]])
map_int(fastq_score_list, count_q)
fastq_df <- fastq_df %>%
              mutate(num_q=map_int(score_list, count_q))

ggplot(fastq_df, aes(x=num_q)) +
  geom_histogram()
```

